Gobba, a purely functional, didactical programming language\\WORK IN PROGRESS DRAFT

### 2020-01-17 paper/paper.tex
fixed typo
renamed project

\title{Minicaml, a purely functional, didactical programming language\\WORK IN PROGRESS DRAFT}
\author{Alessandro Cheli\\Course taught by Prof. Gianluigi Ferrari\\and Prof. Francesca Levi}

@@

\begin{abstract}
\textbf{minicaml} is a dyhamically typed and purely functional interpreted
programming language. It is based on the Professor Gianluigi Ferrari and
Professor Francesca Levi's minicaml, an evaluation example to show students
attending the Programming 2 course at the University of Pisa how interpreters
work. It is an interpreted language heavily inspired from the OCaml, Haskell and
Scheme languages, with static (lexical scoping), eager and lazy evaluation and a
didactical REPL that shows each AST expression and each evaluation step.
\end{abstract}

\section{REPL and command line interface}
\subsection{Installation}
\textbf{minicaml} is available in the opam 2.0 repository.
(\url{https://opam.ocaml.org/}). The easiest way to install minicaml is with the
OCaml package manager \textbf{opam}. To do so, please check that you have a version of opam $\geq$
2.0.0 and run:
\begin{lstlisting}[style=bash]
opam install minicaml
\end{lstlisting}
Alternatively, \textbf{minicaml} can be installed from source by downloading the
source code repository and building it manually. \textbf{minicaml} has been tested
only on Linux and macOS systems. It has not been tested yet on Windows and BSD
derived systems.
\begin{lstlisting}[style=bash]
# download the source code
git clone https://github.com/0x0f0f0f/minicaml
# cd into the source code directory
cd minicaml
# install dependencies
opam install ANSITerminal dune ppx_deriving menhir \\
  cmdliner alcotest bisect_ppx ocamline
# compile
make
# test
make test
# execute
make run
# install
make install
\end{lstlisting}

@@

\section{Purity Inference}
An important feature of the minicaml language is the purity inference algorithm,
which is performed statically on expressions before evaluation. It is an interpretation of expressions over the domain of purity, meant to
prevent side effects by signal an error if they are contained inside the
programs written in the language. Expressions are tagged by the algorithm with
the \texttt{Pure}, \texttt{Impure} and \texttt{Numerical} labels. An
\texttt{Impure} expression is an expression that contains calls to primitives
that perform I/O operations, mutable variables and/or imperative style
assignments. A \texttt{Numerical} expression is an expression where only
numerical operations are performed; \texttt{Pure} expressions are those which do not fall into the previous two
categories.

@@

\section{Evaluator}
\texttt{minicaml}'s evaluator is heavily inspired by the Metacircular Evaluator defined in the
highly acclaimed textbook \textit{Structure and Interpretation of Computer Programs} \cite{Abelson1996}.

\section{Primitives}
The language primitives that are implemented in OCaml are organized in modules
separated by functionality. Each primitive is a function that accepts a list of
evaluated values and returns a single reduced value; therefore they have a type
of \texttt{evt list -> evt}. OCaml primitives have to perform internal
typechecking and unpacking of the arguments they receive from the minicaml
calls.

@@

Some primitives, such as catamorphic procedures, are not native OCaml functions
but small expressions written directly in minicaml; those primitives are kept as
lazy expressions into the same table as native OCaml primitives. The key
difference between the two resides in the fact that those textual minicaml
primitives are not transformed into a function which body contains only an
\texttt{ApplyPrimitive} call, but are instead parsed and analyzed at run time.
The resulting additional startup time caused by parsing and analysis is
proportional to the number of textual form primitives in the table and therefore
quite irrelevant on non-embedded computer systems. The \textit{fold left} and
\textit{fold right} catamorphic primitives are written directly in the minicaml
language and are hereby provided as examples.

>>>>

\title{Gobba, a purely functional, didactical programming language\\WORK IN PROGRESS DRAFT}
\author{Alessandro Cheli\\Course taught by Prof. Gianluigi Ferrari\\and Prof. Francesca Levi}

@@

\begin{abstract}
\textbf{gobba} is a dynamically typed and purely functional interpreted
programming language. It is based on the Professor Gianluigi Ferrari and
Professor Francesca Levi's minicaml, an evaluation example to show students
attending the Programming 2 course at the University of Pisa how interpreters
work. It is an interpreted language heavily inspired from the OCaml, Haskell and
Scheme languages, with static (lexical scoping), eager and lazy evaluation and a
didactical REPL that shows each AST expression and each evaluation step.
\end{abstract}

\section{REPL and command line interface}
\subsection{Installation}
\textbf{gobba} is available in the opam 2.0 repository.
(\url{https://opam.ocaml.org/}). The easiest way to install gobba is with the
OCaml package manager \textbf{opam}. To do so, please check that you have a version of opam $\geq$
2.0.0 and run:
\begin{lstlisting}[style=bash]
opam install gobba
\end{lstlisting}
Alternatively, \textbf{gobba} can be installed from source by downloading the
source code repository and building it manually. \textbf{gobba} has been tested
only on Linux and macOS systems. It has not been tested yet on Windows and BSD
derived systems.
\begin{lstlisting}[style=bash]
# download the source code
git clone https://github.com/0x0f0f0f/gobba
# cd into the source code directory
cd gobba
# install dependencies
opam install ANSITerminal dune ppx_deriving menhir \\
  cmdliner alcotest bisect_ppx ocamline
# compile
make
# test
make test
# execute
make run
# install
make install
\end{lstlisting}

@@

\section{Purity Inference}
An important feature of the gobba language is the purity inference algorithm,
which is performed statically on expressions before evaluation. It is an interpretation of expressions over the domain of purity, meant to
prevent side effects by signal an error if they are contained inside the
programs written in the language. Expressions are tagged by the algorithm with
the \texttt{Pure}, \texttt{Impure} and \texttt{Numerical} labels. An
\texttt{Impure} expression is an expression that contains calls to primitives
that perform I/O operations, mutable variables and/or imperative style
assignments. A \texttt{Numerical} expression is an expression where only
numerical operations are performed; \texttt{Pure} expressions are those which do not fall into the previous two
categories.

@@

\section{Evaluator}
\texttt{gobba}'s evaluator is heavily inspired by the Metacircular Evaluator defined in the
highly acclaimed textbook \textit{Structure and Interpretation of Computer Programs} \cite{Abelson1996}.

\section{Primitives}
The language primitives that are implemented in OCaml are organized in modules
separated by functionality. Each primitive is a function that accepts a list of
evaluated values and returns a single reduced value; therefore they have a type
of \texttt{evt list -> evt}. OCaml primitives have to perform internal
typechecking and unpacking of the arguments they receive from the gobba
calls.

@@

Some primitives, such as catamorphic procedures, are not native OCaml functions
but small expressions written directly in gobba; those primitives are kept as
lazy expressions into the same table as native OCaml primitives. The key
difference between the two resides in the fact that those textual gobba
primitives are not transformed into a function which body contains only an
\texttt{ApplyPrimitive} call, but are instead parsed and analyzed at run time.
The resulting additional startup time caused by parsing and analysis is
proportional to the number of textual form primitives in the table and therefore
quite irrelevant on non-embedded computer systems. The \textit{fold left} and
\textit{fold right} catamorphic primitives are written directly in the gobba
language and are hereby provided as examples.
